Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2014-06-13 Referee1_Response.tex

\begin{itemize}
\item[4.] \textit{p3, l38: Suggest: estimator of the optimality gap \ldots}
\end{itemize}

\noindent 
Done. 
\medskip

\begin{itemize}
\item[5.] \textit{p4, l34: Suggest: once the data are generated through a simulation, they \ldots}
\end{itemize}

\noindent 
Done. 
%Please note that this sentence has been changed slightly based on the comments of the other referee. 
\medskip

@@

\begin{itemize}
\item[6.] \textit{p4, l42: Suggest: Partial overlap results in fewer batches, hence fewer optimization \ldots}
\end{itemize}

\noindent 
We reworded this sentence to read as ``Partial overlapping results in a smaller number of batches; hence, fewer optimization \ldots '' based on the comments of the other referee.
\medskip

@@

\begin{itemize}
\item[7.] \textit{p6, l32: Suggest: The degrees \ldots make \ldots}
\end{itemize}

\noindent 
Done. 
% This sentence might be deleted or changed to the denominator term, etc. CHECK! 
\medskip

@@

\begin{itemize}
\item[16.] \textit{p17, l55: Perhaps add a comment on why DB1 may not experience the asymptotic variance reduction? 
(Is it because it may lack a unique solution?)}
\end{itemize}

\noindent
Done.
% xxx Yes, it is partially because we are not sure if it has a unique optimal solution as DB1 has never been solved before. 
%Even if it has been solved, the variance reduction is shown asymptotically but the computations are only for small sample sizes. 
%We reworded this to reflect these points.
%Thank you for this suggestion. 
\medskip

\begin{itemize}
\item[17.] \textit{p18, figures: The figures are a bit small and hence can be hard to read without enlarging them (but perhaps this will be taken care of at production).}
\end{itemize}

\noindent 
We split these two figures into three and enlarged each graph. 
All the new graphs added to the paper in Section 5.4 (Computational Efficiency) are also larger compared to those in the previous version of the paper. 
\medskip

@@

\begin{itemize}
\item[18.] \textit{p18, l57: The authors speak of coverage probabilities for DB1, but how can this be done without knowing an optimal solution to DB1?}
\end{itemize}

\noindent 
We used a Latin hypercube sample of size 50,000 estimate $\mu_{\hat{x}}$ for DB1, which was then used to estimate the coverage probability of the OMRP CIs for this test problem.
We explain this at the end of Section 5.1 (Test Problems). 
We also changed the wording in Section 5.3 (Results of Experiments) to make it clear that the coverage probability is estimated for DB1. 
\medskip

\begin{itemize}
\item[19.] \textit{p19: It may be possible to expand on the last paragraph of Section 5, in giving guidance on selecting the degree of overlap, depending on the authors' response to comment M1 above.}
\end{itemize}

\noindent 
Thank you for this suggestion. 
xxx We expanded this paragraph based on our findings. 
xxx Please note that this paragraph now appears in a new section (\S 5.4 Computational Efficiency).

### 2014-06-13 omrp_lb_rev.tex
first pass through the referee minor suggestions
more to be done...

A natural setting where batch means are used in stochastic optimization (especially in stochastic programming) is for assessing solution quality. 
This typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonovelapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators applying the classical nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an ovelapping batches version of MRP, allowing for partial overlap, and provide conditions under which the resulting gap estimators exhibit reduced variance when compared to their counterparts obtained via nonoverlapping batches.
We also show that the resulting point estimators are strongly consistent and the interval estimators are asymptotically valid. 
We also present computational results to examine small-sample behavior. 
Our experiments indicate that the use of overlapping estimators results in gap estimators achieving lower variance than their nonoverlapping counterparts, along with comparable bias and interval coverage---for both small- and large-sample cases.

@@

The optimality gap $\e{f(\xh,\xit)} - \zs$ often cannot be calculated exactly. First, for a given candidate solution $\xh \in X$, evaluation of $\e{f(\xh,\xit)}$ typically involves a difficult multidimensional integral. 
Second, we do not know $\zs$. 
In many fields of optimization, the second problem is alleviated by obtaining lower bounds on $\zs$ through relaxations such as integrality, Lagrangian, or semidefinite relations.
As a result, it is common to evaluate an upper bound on the optimality gap. 
This motivates estimating the upper bound on the optimality gap in stochastic programming by using Monte Carlo simulation.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $$
\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}
$$ 
due to the inequality $\e{\zs_m} \leq \zs$ \citep{Mak1999,norkin_pflug_ruszczynski_98}.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, we can simply use $\zs_m$.  
The resulting point estimator of the optimality gap is then 
$$
\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m.
$$ 
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms.

@@

Our aim is to apply the idea of overlapping batches to MRP.  
Overlapping could help to obtain better variance estimators with simple data reuse, resulting in improved CIs.
%Especially for problems where sampling error is dominant, this simple data reuse method could yield better estimators.
We note several differences in this setting compared to the classical setting.  
In simulation output analysis, the point of interest is that of estimating the variance of the sample mean of a covariance stationary process.  
We are interested in estimating the variance of an optimality gap estimator.  
Notice that an optimality gap estimator not only has a sample mean $(\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti))$ but also a minimized sample mean $(\zs_m)$.  
Minimization changes the statistical properties of sample means.  
For instance, the central limit theorem may not hold for a minimized sample mean, even though it holds for each $\x \in \X$.  
Therefore, analysis of the asymptotic benefits of overlapping becomes more complicated.
We overcome this technical difficulty by approximating the optimality gap estimators by their nonoptimized counterparts (see \S \ref{sec:theory}).  
The nonoptimized counterparts are (regular) overlapping batch means estimators (without any optimization present); therefore, they have the desired statistical properties of approximately the same bias and lower variances compared to their nonoverlapping versions. 
We establish convergence of the optimality gap estimators that use overlapping to their nonoptimized counterparts. 
Another difference between our setting and simulation output analysis is that once the data are generated through a simulation, it can be reused without much additional computational effort to obtain the OBM variance estimator.  
In this setting, due to the need to solve a sampling problem (SP$_m$), the computational effort may increase with data reuse.  
This can be alleviated in two ways. 
First, near-optimal variance reduction can be obtained by partially overlapping the batches \citep{Welch1987,Song1992}.  
Partial overlapping results in a smaller number of batches; hence, fewer optimization problems need to be solved.
In this paper, we consider variable amounts of overlap and examine this both theoretically and empirically. 
Second, in many solution methods, warm-starting can be used to quickly solve the sampling approximations with overlapping samples, considerably reducing solution time.

@@

The sample variance estimator is updated to
\begin{equation} \label{eq:overlap_variance_def}
	\vart{\ybb} = \frac{1}{(\tfrac{n}{m}-1)(n - m + 1)}\sum_{j=1}^{n-m+1} (\yb_j(m) - \ybb)^2.
\end{equation}
Note that the corresponding overlapping batch means are no longer independent. 
However, by simply reusing the data in this fashion---and seemingly in a contradictory way that induces dependence---one obtains a better variance estimator. 
\citet{Meketon1984} show that: (i) the overlapping variance estimator has nearly the same bias as the standard nonoverlapping variance estimator, and (ii) the overlapping estimator has only two-thirds of the asymptotic variance of the nonoverlapping one.  
That is, 
\begin{equation} \label{eq:asym_var}
	\frac{ \var{\vart{\ybb}} }{ \var{\varhat(\ybb)} } \rightarrow \frac{2}{3}
\end{equation}
in the limit as batch size $m$ and the number of batches $(n/m)$ tend to infinity.  
Note that the degrees of freedom in (\ref{eq:overlap_variance_def}) is slightly different than in \citep{Meketon1984}.  
The degrees of freedom in (\ref{eq:overlap_variance_def}) makes $\vart{\ybb}$ an unbiased estimator for i.i.d.\ data for all finite $m$ and $n$ with the same asymptotic benefits \citep{Song1992}.  
The $(1-\alpha)$-level approximate CI is similarly formed by $\ybb \pm t_{3(k-1)/2,\alpha/2} \sqrt{\vart{\ybb}}$, with an adjustment to the degrees of freedom.

@@

Taking $n$ observations of the random variables $\xit$ and splitting them into $k$ nonoverlapping batches of size $m$, define 
\begin{align} 
	\gb_j  & = \frac{1}{m} \sum_{i = 1}^{m} f(\xh, \xit^{m(j-1)+i}) - \min_{\x \in X} \frac{1}{m} \sum_{i=1}^{m} f(\x, \xit^{m(j-1)+i}), \ \ \ \ j = 1,2, \dots, k \label{eq:mrp_gap}\\
         & = \frac{1}{m} \sum_{i = 1}^{m} \left[ f(\xh, \xit^{m(j-1)+i}) -  f(\xs_j, \xit^{m(j-1)+i}) \right], \ \ \ \ j = 1,2, \dots, k, \notag
\end{align}
where $\xs_j$ denotes an optimal solution to (\ref{eq:sto_prog_m}) of batch $j$ given on the right-hand side of (\ref{eq:mrp_gap}).    
Since we are assessing the quality of a given solution $\xh \in \X$, we suppress it from the notation $\gb_j$.
The optimality gap estimator $\gb_j$ is similar to $\yb_j$ with individual observations $y^{m(j-1)+i} = f(\xh,\xit^{m(j-1)+i}) - f(\xs_j,\xit^{m(j-1)+i})$, for $i=1,\ldots,m$ and $j=1,\ldots, k$.
Notice that, in this case, $y^i$ not only depends on observation $i$ of batch $j$, $\xit^{m(j-1)+i}$, but all the observations in that batch $\xit^{m(j-1)+1},\xit^{m(j-1)+2},\ldots,\xit^{m(j-1)+m}$ because $\xs_j$ is the optimal solution to (\ref{eq:sto_prog_m}) given in (\ref{eq:mrp_gap}).   
As before, after $k$ estimates are obtained, the overall mean of these estimates $\gbb = \frac{1}{k} \sum_{j=1}^k \gb_j$ provides a point estimator of the optimality gap.  
The sample variance is obtained as in (\ref{eq:var}) by $\varh{\gbb} = \frac{1}{k} \frac{1}{k-1} \sum_{j=1}^k (\gb_j - \gbb)^2$, which results in a one-sided CI $\left[0, \gbb + t_{k-1,\alpha} \sqrt{\varh{\gbb}} \right]$ that has a level of significance, which is approximately $\alpha$.

@@

\subsection{Definition of Estimators for OMRP}
In order to apply overlapping batches to MRP, we need to keep track of solutions to sampling problems (SP$_m$) for each batch of size $m$.  
Toward this end, let $B(i)$ denote the set of batches $j \in \left\{1, 2, \dots, \nb \right\}$ observation $\xiti$ is used in, $i = 1, 2, \dots, n$.  
See Figure \ref{fig:overlap_nonint} for an example with $n=18$, $m = 6$ and $\gamma = 2$.  
Here, the first batch uses observations $\xit^1, \xit^2, \xit^3, \xit^4, \xit^5, \xit^6$, the second batch uses $\xit^3, \xit^4, \xit^5, \xit^6, \xit^7, \xit^8$, and so on.  
So, $B(1) = \{1\}$, $B(3) = \{1,2\}$, and $B(7)=\{2,3,4\}$.  
As before, we use $\xs_j$ to denote an optimal solution to sampling problem (SP$_m$) formed using the $j$th batch; $\xs_j \in \argmin_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xit^{\gamma(j-1) + i})$, $j = 1, 2, \dots, \nb$.

@@

First, we restrict our attention to i.i.d.\ sampling by A1.
The classical variance-reduction results for overlapping batches show that $\var{n_l VG_l}$ is reduced by overlapping, thus requiring at least finite fourth moments. 
Finally, our proof relies on an exponential rate of convergence of sampled solution to the true solution of the problem.  
Assumption A3 can be satisfied for different classes of problems with unique optimal solutions.  
\citet{shapiro2000rate} establish exponential rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite.   
\citet{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

	The fringe batches will come in identically distributed pairs.  
        The first and last batch would be identically distributed, as would the second and second-to-last, etc.  
        We know this because we could reverse the order of the i.i.d.\ sample and perform the batching again to get the exact same batch structure.  
        Rewriting to take this into account, letting $A_j = (\gb_j - \gbb)^2 - (\db_j - \dbb)^2$, and continuing from (\ref{eq:vvar_conv_pf_1}),
	\begin{align*}
		& \left( \frac{n}{dt} \right)^2 n_b \e{ \sum_{1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \\
		 =  & \left( \frac{n}{dt} \right)^2 n_b \e{\left[ \sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=n_b-\left\lceil \gammab^{-1} \right\rceil+1}^{n_b} \left[ A_j \right]^2 \right]\
		} \\
		 = & \left( \frac{n}{dt} \right)^2 n_b \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } + \left( \frac{n}{dt} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } \\
		 = & \left( \frac{n}{dt} \right)^2 n_b \left(n_b - 2\left\lceil \gammab^{-1} \right\rceil \right) \e{ \left[ A_j \right]^2 }
		+ \left( \frac{n}{dt} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 }.
	\end{align*}	
	We have $\left\lceil \gammab^{-1} \right\rceil + 1$ expectation terms, each of which has a coefficient that is less than $\left(\frac{n}{dt}\right)^2n_b^2 < (2m)^2$.  
        So each expectation can be bounded by
	\begin{align}
		\left( 2m \right)^2 \e{ \left[ A_j \right]^2 } & = \left( 2m \right)^2 \e{ \left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^2 \left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^2 } \notag \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{\e{\left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^4 }} \label{eq:var_reduct_cauchy2} \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{8 \left(\e{(\gb_j - \gbb)^4} + \e{(\db_j - \dbb)^4} \right)}, \label{eq:var_reduct_holder2}
	\end{align}
	where equation (\ref{eq:var_reduct_cauchy2}) follows from the Cauchy-Schwartz inequality, and equation (\ref{eq:var_reduct_holder2}) follows from H\"{o}lder's inequality.  
        Since we can bound both expectation terms in (\ref{eq:var_reduct_holder2}) by A2, we can demonstrate $\sqrt{\frac{n}{m}}(nVG - nVD) \xrightarrow{L^2} 0$ by focusing on only one of the terms.  
        Focusing on the first expectation term in (\ref{eq:var_reduct_holder2}) and returning the $\tfrac{n}{m}$, by H\"{o}lder's inequality,
	\begin{align*}
		\frac{n}{m} 4m^2 \e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}	& \leq 32 \frac{n}{m} m^2\left( \e{(\db_j - \gb_j)^4} + \e{(\dbb - \gbb)^4} \right).
	\end{align*}	
	By Lemma \ref{lem:gb_gbb_l4}, both expectation terms above converge to zero with exponential rates.  
        With a batching structure $m_l = \lfloor n_l^r \rfloor$ for some $0<r<1$, we see that $\frac{n_l}{m_l} \e{(n_l VG_l - n_l VD_l)^2} \rightarrow 0$.
\end{proof}

@@

The asymptotic validity of the OMRP CI requires $n_l VD_l$ to converge in probability to $\sigma^{2}_{\xh}$. 
This can be achieved either by almost sure or mean-square convergence, which gives a larger range of $r$ in the batching structure.
Also, a hypothesis of Theorem~\ref{thm:conf_int} is that $\xh \neq \xs$.  
For $\xh = \xs$, $\mu_\xh = 0$, and therefore, we have $\lim_{l\rightarrow\infty} \p{\mu_\xh \leq I^G_l} = 1$.

@@

A summary of results from these experiments is depicted in Figures~\ref{fig:varvar1}--\ref{fig:nv}.
Figures~\ref{fig:varvar1} and \ref{fig:varvar2} show the reduction of variance, with each term being normalized with respect to $\var{nVG}$ of the nonoverlapping case (MRP). 
The solid lines in these figures show the theoretical variance reduction from \citep{Welch1987} given by (\ref{eq:var_reduct_formula}). 
Empirical results agree with the theoretical values well, with more variability observed in CEP1, 10D, and DB1. 
These experiments suggest that similar variance reduction can be achieved in stochastic programming by overlapping the batches, even with small sample sizes.
We note that the variance reduction in DB1 can be less than the asymptotic amount. %for the small sample sizes used in these experiments. 
Due its size, DB1 has not yet been solved to optimality, therefore, it may lack a unique optimal solution. 
Even with the relatively small sample sizes compared to the size of this problem, however, overlapping appears to achieve some variance reduction.

>>>>

A natural setting where batch means are used in stochastic optimization (especially in stochastic programming) is for assessing solution quality. 
This typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonoverlapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators applying the classical nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an overlapping batches version of MRP, allowing for partial overlap, and provide conditions under which the resulting gap estimators exhibit reduced variance when compared to their counterparts obtained via nonoverlapping batches.
We also show that the resulting point estimators are strongly consistent and the interval estimators are asymptotically valid. 
We also present computational results to examine small-sample behavior. 
Our experiments indicate that the use of overlapping estimators results in gap estimators achieving lower variance than their nonoverlapping counterparts, along with comparable bias and interval coverage---for both small- and large-sample cases.

@@

The optimality gap $\e{f(\xh,\xit)} - \zs$ often cannot be calculated exactly. First, for a given candidate solution $\xh \in X$, evaluation of $\e{f(\xh,\xit)}$ typically involves a difficult multidimensional integral. 
Second, we do not know $\zs$. 
In many fields of optimization, the second problem is alleviated by obtaining lower bounds on $\zs$ through relaxations such as integrality, Lagrangian, or semidefinite relaxations.
As a result, it is common to evaluate an upper bound on the optimality gap. 
This motivates estimating the upper bound on the optimality gap in stochastic programming by using Monte Carlo simulation.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $$
\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}
$$ 
due to the inequality $\e{\zs_m} \leq \zs$ \citep{Mak1999,norkin_pflug_ruszczynski_98}.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, we can simply use $\zs_m$.  
The resulting point estimator of the optimality gap is then 
$$
\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m.
$$ 
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms.

@@

Our aim is to apply the idea of overlapping batches to MRP.  
Overlapping could help to obtain better variance estimators with simple data reuse, resulting in improved CIs.
%Especially for problems where sampling error is dominant, this simple data reuse method could yield better estimators.
We note several differences in this setting compared to the classical setting.  
In simulation output analysis, the point of interest is that of estimating the variance of the sample mean of a covariance stationary process.  
We are interested in estimating the variance of an optimality gap estimator.  
Notice that an optimality gap estimator not only has a sample mean $(\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti))$ but also a minimized sample mean $(\zs_m)$.  
Minimization changes the statistical properties of sample means.  
For instance, the central limit theorem may not hold for a minimized sample mean, even though it holds for each $\x \in \X$.  
Therefore, analysis of the asymptotic benefits of overlapping becomes more complicated.
We overcome this technical difficulty by approximating the optimality gap estimators by their nonoptimized counterparts (see \S \ref{sec:theory}).  
The nonoptimized counterparts are (regular) overlapping batch means estimators (without any optimization present); therefore, they have the desired statistical properties of approximately the same bias and lower variances compared to their nonoverlapping versions. 
We establish convergence of the optimality gap estimators that use overlapping to their nonoptimized counterparts. 
Another difference between our setting and simulation output analysis is that once the data are generated through a simulation, they can be reused without much additional computational effort to obtain the OBM variance estimator.  
In this setting, due to the need to solve a sampling problem (SP$_m$), the computational effort may increase with data reuse.  
This can be alleviated in two ways. 
First, near-optimal variance reduction can be obtained by partially overlapping the batches \citep{Welch1987,Song1992}.  
Partial overlapping results in a smaller number of batches; hence, fewer optimization problems need to be solved.
In this paper, we consider variable amounts of overlap and examine this both theoretically and empirically. 
Second, in many solution methods, warm-starting can be used to quickly solve the sampling approximations with overlapping samples, considerably reducing solution time.

@@

The sample variance estimator is updated to
\begin{equation} \label{eq:overlap_variance_def}
	\vart{\ybb} = \frac{1}{(\tfrac{n}{m}-1)(n - m + 1)}\sum_{j=1}^{n-m+1} (\yb_j(m) - \ybb)^2.
\end{equation}
Note that the corresponding overlapping batch means are no longer independent. 
However, by simply reusing the data in this fashion---and seemingly in a contradictory way that induces dependence---one obtains a better variance estimator. 
\citet{Meketon1984} show that: (i) the overlapping variance estimator has nearly the same bias as the standard nonoverlapping variance estimator, and (ii) the overlapping estimator has only two-thirds of the asymptotic variance of the nonoverlapping one.  
That is, 
\begin{equation} \label{eq:asym_var}
	\frac{ \var{\vart{\ybb}} }{ \var{\varhat(\ybb)} } \rightarrow \frac{2}{3}
\end{equation}
in the limit as batch size $m$ and the number of batches $(n/m)$ tend to infinity.  
Note that the degrees of freedom in (\ref{eq:overlap_variance_def}) is slightly different than in \citep{Meketon1984}.  
The degrees of freedom in (\ref{eq:overlap_variance_def}) make $\vart{\ybb}$ an unbiased estimator for i.i.d.\ data for all finite $m$ and $n$ with the same asymptotic benefits \citep{Song1992}.  
The $(1-\alpha)$-level approximate CI is similarly formed by $\ybb \pm t_{3(k-1)/2,\alpha/2} \sqrt{\vart{\ybb}}$, with an adjustment to the degrees of freedom.

@@

Taking $n$ observations of the random variables $\xit$ and splitting them into $k$ nonoverlapping batches of size $m$, define 
\begin{align} 
	\gb_j  & = \frac{1}{m} \sum_{i = 1}^{m} f(\xh, \xit^{m(j-1)+i}) - \min_{\x \in X} \frac{1}{m} \sum_{i=1}^{m} f(\x, \xit^{m(j-1)+i}), \ \ \ \ j = 1,2, \dots, k \label{eq:mrp_gap}\\
         & = \frac{1}{m} \sum_{i = 1}^{m} \left[ f(\xh, \xit^{m(j-1)+i}) -  f(\xs_j, \xit^{m(j-1)+i}) \right], \ \ \ \ j = 1,2, \dots, k, \notag
\end{align}
where $\xs_j$ denotes an optimal solution to (\ref{eq:sto_prog_m}) of batch $j$ given on the right-hand side of (\ref{eq:mrp_gap}).    
Since we are assessing the quality of a given solution $\xh \in \X$, we suppress it from the notation $\gb_j$.
The optimality gap estimator $\gb_j$ is similar to $\yb_j$ with individual observations $y^{m(j-1)+i} = f(\xh,\xit^{m(j-1)+i}) - f(\xs_j,\xit^{m(j-1)+i})$, for $i=1,\ldots,m$ and $j=1,\ldots, k$.
Notice that, in this case, $y^i$ not only depends on observation $i$ of batch $j$, $\xit^{m(j-1)+i}$, but all the observations in that batch $\xit^{m(j-1)+1},\xit^{m(j-1)+2},\ldots,\xit^{m(j-1)+m}$ because $\xs_j$ is the optimal solution to (\ref{eq:sto_prog_m}) given in (\ref{eq:mrp_gap}).   
As before, after $k$ estimates are obtained, the overall mean of these estimates $\gbb = \frac{1}{k} \sum_{j=1}^k \gb_j$ provides a point estimator of the optimality gap.  
The sample variance is obtained as in (\ref{eq:var}) by $\varh{\gbb} = \frac{1}{k} \frac{1}{k-1} \sum_{j=1}^k (\gb_j - \gbb)^2$, which results in an approximate $(1-\alpha)$-level one-sided CI $\left[0, \gbb + t_{k-1,\alpha} \sqrt{\varh{\gbb}} \right]$.

@@

\subsection{Definition of Estimators for OMRP}

In order to apply overlapping batches to MRP, we need to keep track of solutions to sampling problems (SP$_m$) for each batch of size $m$.  
Toward this end, let $B(i)$ denote the set of batches $j \in \left\{1, 2, \dots, \nb \right\}$ observation $\xiti$ is used in, $i = 1, 2, \dots, n$.  
See Figure \ref{fig:overlap_nonint} for an example with $n=18$, $m = 6$ and $\gamma = 2$.  
Here, the first batch uses observations $\xit^1, \xit^2, \xit^3, \xit^4, \xit^5, \xit^6$, the second batch uses $\xit^3, \xit^4, \xit^5, \xit^6, \xit^7, \xit^8$, and so on.  
So, $B(1) = \{1\}$, $B(3) = \{1,2\}$, and $B(7)=\{2,3,4\}$.  
As before, we use $\xs_j$ to denote an optimal solution to sampling problem (SP$_m$) formed using the $j$th batch; $\xs_j \in \argmin_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xit^{\gamma(j-1) + i})$, $j = 1, 2, \dots, \nb$.

@@

First, we restrict our attention to i.i.d.\ sampling by A1.
The classical variance-reduction results for overlapping batches show that $\var{n_l VG_l}$ is reduced by overlapping, thus requiring at least finite fourth moments. 
Finally, our proof relies on an exponential rate of convergence of the sampled solution to the true solution of the problem.  
Assumption A3 can be satisfied for different classes of problems with unique optimal solutions.  
\citet{shapiro2000rate} establish exponential rates of convergence for a class of (\ref{eq:sto_prog}), which include two-stage stochastic linear programs with recourse when the support of $\xit$ is finite.   
\citet{kleywegt2002sample} show that assumption A3 can be satisfied for discrete stochastic programs (i.e., when $X$ is a finite set) under the condition that the moment generating function of $f(\xs,\xit) - f(\x,\xit)$ for $x \in X \backslash \{\xs\}$ is finite valued on $\R$.  
This class of problems includes stochastic integer programs with recourse.
  We present computational results on both stochastic linear and integer programs with recourse in \S \ref{sec:comp}.

@@

	The fringe batches come in identically distributed pairs.  
        The first and last batch are identically distributed, as are the second and second-to-last, etc.  
        We know this because we could reverse the order of the i.i.d.\ sample and perform the batching again to get the exact same batch structure.  
        Rewriting to take this into account, letting $A_j = (\gb_j - \gbb)^2 - (\db_j - \dbb)^2$, and continuing from (\ref{eq:vvar_conv_pf_1}),
	\begin{align*}
		& \left( \frac{n}{dt} \right)^2 n_b \e{ \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \\
		 =  & \left( \frac{n}{dt} \right)^2 n_b \e{\left[ \sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \left[ A_j \right]^2 \
		 + \sum_{j=n_b-\left\lceil \gammab^{-1} \right\rceil+1}^{n_b} \left[ A_j \right]^2 \right]\
		} \\
		 = & \left( \frac{n}{dt} \right)^2 n_b \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } + \left( \frac{n}{dt} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 } \\
		 = & \left( \frac{n}{dt} \right)^2 n_b \left(n_b - 2\left\lceil \gammab^{-1} \right\rceil \right) \e{ \left[ A_j \right]^2 }
		+ \left( \frac{n}{dt} \right)^2 (n_b) 2\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left[ A_j \right]^2 }.
	\end{align*}	
	We have $\left\lceil \gammab^{-1} \right\rceil + 1$ expectation terms, each of which has a coefficient that is less than $\left(\frac{n}{dt}\right)^2n_b^2 < (2m)^2$.  
        So each expectation can be bounded by
	\begin{align}
		\left( 2m \right)^2 \e{ \left[ A_j \right]^2 } & = \left( 2m \right)^2 \e{ \left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^2 \left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^2 } \notag \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{\e{\left[(\gb_j - \gbb) + (\db_j - \dbb)\right]^4 }} \label{eq:var_reduct_cauchy2} \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}} \sqrt{8 \left(\e{(\gb_j - \gbb)^4} + \e{(\db_j - \dbb)^4} \right)}, \label{eq:var_reduct_holder2}
	\end{align}
	where equation (\ref{eq:var_reduct_cauchy2}) follows from the Cauchy-Schwartz inequality, and equation (\ref{eq:var_reduct_holder2}) follows from H\"{o}lder's inequality.  
        Since we can bound both expectation terms in (\ref{eq:var_reduct_holder2}) by A2, we can demonstrate $\sqrt{\frac{n}{m}}(nVG - nVD) \xrightarrow{L^2} 0$ by focusing on only one of the terms.  
        Focusing on the first expectation term in (\ref{eq:var_reduct_holder2}) and returning the $\tfrac{n}{m}$, by H\"{o}lder's inequality,
	\begin{align*}
		\frac{n}{m} 4m^2 \e{\left[(\gb_j - \gbb) - (\db_j - \dbb)\right]^4}	& \leq 32 \frac{n}{m} m^2\left( \e{(\db_j - \gb_j)^4} + \e{(\dbb - \gbb)^4} \right).
	\end{align*}	
	By Lemma \ref{lem:gb_gbb_l4}, both expectation terms above converge to zero with exponential rates.  
        With a batching structure $m_l = \lfloor n_l^r \rfloor$ for some $0<r<1$, we see that $\frac{n_l}{m_l} \e{(n_l VG_l - n_l VD_l)^2} \rightarrow 0$.
\end{proof}

@@

The asymptotic validity of the OMRP CI requires $n_l VD_l$ to converge in probability to $\sigma^{2}_{\xh}$. 
This can be achieved either by almost sure or mean-square convergence, where the latter gives a larger range of $r$ in the batching structure.
Also, a hypothesis of Theorem~\ref{thm:conf_int} is that $\xh \neq \xs$.  
For $\xh = \xs$, $\mu_\xh = 0$; therefore, in this case, we have $\lim_{l\rightarrow\infty} \p{\mu_\xh \leq I^G_l} = 1$.

@@

A summary of results from these experiments is depicted in Figures~\ref{fig:varvar1}--\ref{fig:nv}.
Figures~\ref{fig:varvar1} and \ref{fig:varvar2} show the reduction of variance, with each term being normalized with respect to $\var{nVG}$ of the nonoverlapping case (MRP). 
The solid lines in these figures show the theoretical variance reduction from \citep{Welch1987} given by (\ref{eq:var_reduct_formula}). 
Empirical results agree with the theoretical values well, with more variability observed in CEP1, 10D, and DB1. 
These experiments suggest that similar variance reduction can be achieved in stochastic programming by overlapping the batches, even with small sample sizes.
We note that the variance reduction in DB1 can be less than the asymptotic amount %for the small sample sizes used in these experiments. 
because due its size, DB1 has not yet been solved to optimality; therefore, it may lack a unique optimal solution. 
%, violating assumption A3. (?)
Even with the relatively small sample sizes compared to the size of this problem, however, overlapping appears to achieve some variance reduction.
